### 2018-03-08 paper/background.tex
Finish Excel subsection

\subsection{The venerable \Make}

\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we therefore use a simple and sensible
approximation to a real \Make that you might find on your machine.} was developed
more than 40 years ago to automatically build software libraries and executable
programs from source code. It uses \emph{makefiles} to describe build tasks,
typically referred to as \emph{build rules}, and their dependencies in a simple
text form. For example:

@@

To achieve minimality \Make relies on two main ideas. First, it uses \emph{file
modification time} to detect the files that are dirty: a file is marked dirty if
it was modified after the previous build. Second, it constructs a complete task
dependency graph from the information contained in the makefile and executes
tasks in a \emph{topological order}.

Note that it is not always possible to know the dependencies of a source file
upfront, or \emph{statically}. To demonstrate this, let us first add another
task to the above makefile:

@@

\subsection{\Excel: a build system in disguise}

@@

\Shake

\begin{figure}[h]
\centerline{\includegraphics[scale=0.28]{fig/make-example-cutoff.pdf}}
\caption{An early cut-off example.\label{fig-cutoff}}
\end{figure}

@@

\Bazel

@@

\begin{itemize}
    \item Many build systems, such as the venerable \Make, need to know the
    complete \emph{dependency graph} between tasks at the start of the build
    process. This makes it possible to analyse the graph ahead of time, which
    helps with efficient task scheduling but is fundamentally limited: parts of
    the dependency graph can often be discovered only during the build process,
    i.e. the dependency graph is \emph{dynamic}, not \emph{static}.

@@

    \item Most build systems track changes of inputs and intermediate results,
    executing dependent tasks whenever they change, but some build systems can
    also track changes in the tasks themselves: if a task has changed, the build
    system will execute it. For example, when a cell's formula has changed,
    \Excel will recompute its value and propagate the changes. We call this
    \emph{self-tracking}. Self-tracking is uncommon in software build systems,
    where one often needs to manually initiate a rebuild of all tasks even if
    just a single task has changed.

@@

\emph{Keys} are used to distinguish \emph{values}. In build systems keys are
typically filenames, e.g. \textsf{src/file.c}, whereas values are file contents
(a C program source code in this case). In spreadsheets keys are cell names,
e.g. \textsf{A1}, and values are numbers, text, etc. that are typically displayed
inside cells. We will use type variables \hs{k} and \hs{v} to denote keys and
values, respectively.

>>>>

\subsection{The venerable \Make: static dependencies and file modification times}

\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we therefore use a simple and sensible
approximation to a real \Make that you might find on your machine.} was developed
more than 40 years ago to automatically build software libraries and executable
programs from source code. It uses \emph{makefiles} to describe build tasks and
their dependencies in a simple text form. For example:

@@

To achieve minimality \Make relies on two main ideas. First, it uses \emph{file
modification time} to detect the files that are dirty: a file is marked dirty if
it was modified since the previous build. Second, it constructs a complete task
dependency graph from the information contained in the makefile and executes
tasks in a \emph{topological order}.

Note that it is not always possible to know the dependencies of a task upfront,
or \emph{statically}. To demonstrate this, let us add another task to the above
makefile:

@@

\subsection{\Excel: dynamic dependencies at the cost of minimality}

\Excel is a build system that can deal with dynamic dependencies albeit at the
cost of sacrificing minimality. Consider the following spreadsheet comprising
four input cells \cmd{A1-A3} and \cmd{B1}, and a single task computing the sum
of values \cmd{A1} $+\cdots+$ \cmd{An}, where \cmd{n} is specified indirectly in
\cmd{B1}:

\begin{minted}[frame=single]{bash}
A1 = 10
A2 = 20
A3 = 30
B1 = 2
C1 = SUM(A1:INDIRECT("A" & B1))
\end{minted}

The output \cmd{C1} here is analogous to the \cmd{release.tar} from the previous
example: one cannot statically determine the dependencies of the associated task.
To deal with this, \Excel takes a simple conservative approach: it marks all
cells with dynamic dependencies as dirty and therefore always recomputes them
during a rebuild. This guarantees that the build result is correct, but clearly
violates the minimality requirement: if the user modifies \cmd{A3}, \Excel will
recompute \cmd{C1} even though it does not transitively depend on \cmd{A3}.

Note that \Excel's approach to marking cells as dirty is a relatively minor
variation of the approach used by \Make: a cell is considered dirty if it was
modified since the previous build \emph{or if it is volatile}, where the notion
of volatility includes dynamic dependencies, as well as \emph{non-determinism}
that will be discussed in~\S\ref{sec-engineering}. We therefore put both \Make
and \Excel in the same category of build systems that associate a \emph{dirty bit}
with every file or cell. (To avoid the ``file or cell'' nuisance, we introduce the
abstract notion of \emph{keys} in~\S\ref{sec-vocabulary}.)

\Excel's recalculation algorithm~\cite{excel_recalc} on the other hand is
significantly different from \Make. Since it is impossible to construct the full
dependency graph upfront in the presence of dynamic dependencies, \Excel
maintains the \emph{calculation chain}, which is an approximation to the correct
topological order. During recalculation, \Excel processes cells in this order,
but can \emph{defer recalculation of a cell} by moving it down the chain when newly
discovered dynamic dependencies dictate that. We discuss this algorithm in more
detail in~\S\ref{sec-examples}.

Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
systems track changes of inputs and intermediate results, executing dependent
tasks whenever they change, but \Excel can also track changes in the tasks
themselves: if a formula was modified, \Excel will recompute it and propagate
the changes. Self-tracking is uncommon in software build systems, where one
often needs to manually initiate a full rebuild even if just a single task has
changed.

\subsection{\Shake: dynamic dependencies with no remorse}

\begin{figure}[h]
\centerline{\includegraphics[scale=0.28]{fig/make-example-cutoff.pdf}}
\caption{An early cutoff example.\label{fig-cutoff}}
\end{figure}

@@

\subsection{\Bazel: a cloud build system}

@@

\subsection{Abstract vocabulary for build systems}\label{sec-vocabulary}

@@

\begin{itemize}

@@

@@

Keys are
used to distinguish values. In build systems keys are typically filenames, e.g.
\cmd{main.c}, whereas values are file contents (a C program source code in this
case). In spreadsheets keys are cell names, e.g. \cmd{A1}, and values are
numbers, text, etc. that are typically displayed inside cells. In Haskell code,
we will use type variables \hs{k} and \hs{v} to denote keys and values,
respectively.
